Financialisation of Commodity Markets \\\large{Co-movement Behind-the-Scenes}

### 2025-08-19 blind.tex
Deploying to gh-pages from @ bautheac/financialisation-co-movement@fba5d0f4311c2970234a42c3d1515050b6d463c6 🚀

Commodity index swaps, commodity index funds (CIFs), commodity-based
exchange traded funds (ETFs) and notes (ETNs) or commodity linked notes
(CLNs) were amongst the main investment vehicles that allowed
institutional and retail investors to build up commodity exposure in the
early 2000s
\citep{boons_stock_2012, henderson_new_2015, irwin_index_2011, unctad_global_2009, schumann_hunger_2011}.
Most of these products were primarily designed to track prominent
commodity indexes among which the S\&P-GSCI, the most popular.\\
Most of its constituents at the time of writing form the basis of the
broad cross-section of commodities that we consider in this study:
corn-\#2 yellow (XCBT), oats (XCBT), soybean meal (XCBT), soybean oil
(XCBT), soybeans (XCBT), wheat-SRW (XCBT), cattle-feeder (XCME),
cattle-live (XCME), lean hogs (XCME), cocoa (IFUS), coffee-C (IFUS),
cotton-\#2 (IFUS), lumber (XCME), orange juice (IFUS), sugar-\#11
(IFUS), natural gas (XNYM), crude oil-WTI (XNYM), gasoline (XNYM),
heating oil (XNYM), copper (XCEC), gold (XCEC), palladium (XNYM),
platinum (XNYM) and silver (XCEC). Unleaded gasoline, the NYMEX legacy
contract for gasoline, stopped trading in 2006 when it was replaced by
Reformulated Gasoline Blendstock for Oxygen Blending (RBOB) with the two
futures series trading alongside for most of the year. For our NYMEX
gasoline data series, we consider unleaded gasoline up to September
1\textsuperscript{st}, 2006 and RBOB gasoline thereafter, date at which
liquidity for the latter overtook that for the former.\\
We also consider the GB metal complex with aluminium (XLME), copper
(XLME), lead (XLME), nickel-primary (XLME), tin-refined (XLME) and zinc
(XLME). This sample cross-section of assets is further divided into
groups for the purpose of the analysis by country (US, GB), sector
(agriculturals, energy, metals) and by sub-sector (agriculturals:
grains, livestock, softs; energy: gas, petroleum; metals: base,
precious)\footnote{See Section~\ref{sec-taxonomy} for a detailed
  taxonomy of the commodity assets considered in the study.}.

@@

  attached
  \href{http://bautheac.duckdns.org:3838/bautheac/co-movement/}{Internet

@@

Financialisation was a global phenomenon, through the channel of index
investment it simultaneously influenced the entire spectrum of liquid
commodities, eventually spurring the transformation of commodity markets
from a collection of idiosyncratic markets into a more cohesive
collection akin to an asset class. In this context, interconnectedness
emerges as a crucial aspect that warrants thorough examination.\\
We strive to address this issue by first examining the average returns
pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector.
Table~\ref{tbl-correlations-inner-periods} shows the results for the
four periods of interest with the results reported for the entire period
as well as for periods of \textbf{backwardation} and \textbf{contango}.
For the entire sample the average pairwise correlation jumps from 0.07
in the first period to 0.16 over financialisation. This finding
indicates higher interdependence within the full cross-section of
commodities starting with the financialisation period, consistent with
the empirical findings in \citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These results thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods. In this context,
the ontological consequences of financialisation are better observed
from an investment perspective as illustrated in
Table~\ref{tbl-correlations-cross-periods} where we observe average
pairwise correlation coefficients between equally weighted portfolios of
US traded commodity assets grouped by sub-sector\footnote{A more
  detailed breakdown of country, sector and subsector average pairwise
  correlations is provided in the attached
  \href{http://bautheac.duckdns.org:3838/bautheac/co-movement/}{Internet
  Appendix}.}. Although modest in magnitude, these exhibit a telling
pattern, mirroring that shown by the individual commodity returns in
Table~\ref{tbl-stats-regime-tests}, with a substantially higher average
coefficient over \textbf{backwardation} phases in the first period (0.12
vs.~0.03) and the pattern reversed over financilisation with an average
coefficient substantially higher in \textbf{contango} (0.12 vs.~0.02).\\
We break down Table~\ref{tbl-correlations-inner-periods}'s average
correlations by year in Table~\ref{tbl-correlations-inner-years}. For
the full sample of assets, the average pairwise correlations remain in
the 0.06--0.08 range from 1998 through 2003, before suddenly increasing
to 0.12 from 2003 to 2004. This suggests 2004 as a distinct break point
for the onset of financialisation, as noted in several studies
\citep{baker_financialization_2021, tang_index_2012}. We further observe
that the US metals appear to have played a major role in this as the
increase seems driven by the sector which average pairwise correlation
more than doubles (0.49 vs 0.22) over the period.

@@

  \href{http://bautheac.duckdns.org:3838/bautheac/co-movement/}{Internet
  Appendix}.}. The onset of financialisation leads to a substantial
improvement in the performance of the CHP factor with its average
\(R^{2}\), albeit still small at 5.4\%, almost tripling from the past to
the financialisation period in contrast to the other factors that show
weaker dynamics with a slight decrease and a much smaller increase for
the open interest and term structure factors respectively. Here again
the increase in pricing performance for the CHP factor is driven by the
metals sector, particularly the precious sub-sector with the average
\(R^{2}\) for the group jumping to almost 26\% during financialisation
from 1.2\% over the past period. The corresponding figures of 22\% and
1.1\% for the entire US metals sector stand in sharp contrast with the
agricultural sector, which exhibits no notable change, and the energy
sector, where pricing performance drops substantially from 9.2\% down to
1.6\%. The improvement in the CHP factor's pricing performance
remarkably comes predominantly from the long leg (low CHP assets:
Keynesian ``normal backwardation'') of the factor\footnote{A breakdown
  of the performance by factor leg is provided in the attached
  \href{http://bautheac.duckdns.org:3838/bautheac/co-movement/}{Internet
  Appendix}.} with the corresponding mimicking portfolio's performance
itself driven by the metals, the precious sub-sector in particular.
Interestingly, although the term structure factor itself only sees a
very modest increase, its short leg (low roll-yield assets: Working's
backwardation) shows a sizeable rise in pricing performance, yet again
driven the metals sector. The pattern of results for both factors
extends to the GB traded assets and suggest a convergence of Keneysian
and Working's term structure perspective over the financialisation
period.\\

@@

\caption{\label{tbl-stats-regime-tests}This table shows mean returns and
volatilities for the individual commodities and two equally weighted
portfolios of US and GB traded assets for the four periods. For each
period, estimates are computed independently over the whole
period--figure shown--as well as over phases of \textbf{backwardation}
(b) and \textbf{contango} (c)--results in brackets. The table further
shows the results of t-tests of significance for the mean returns
computed over the whole periods and t-tests of difference between these
when computed over the two phases independently along with F-tests of
difference for the corresponding variances. Values significant at the
1\%, 5\% and 10\% level are marked with ***, ** and * respectively.
Detailed results are available in the attached
\href{http://bautheac.duckdns.org:3838/bautheac/co-movement/}{Internet
Appendix}. \textbf{Backwardation} and \textbf{contango} regimes as well
as period definitions are discussed in Section~\ref{sec-methods} while
the results are discussed in Section~\ref{sec-results}.}

@@

\caption{\label{tbl-regressions-factors}This table shows the average
time series regression \(R^{2}\) for models where the returns series for
all the commodity assets are independently regressed against the return
series for our mimicking portfolios for risk factors in returns,
including the market, CHP, open interest and term structure factors. The
results are computed independently across countries and commodity
sectors and sub-sectors for the four periods. A breakdown of the
estimates computed independently over the whole periods as well as over
phases of \textbf{backwardation} and \textbf{contango} is provided in
the attached
\href{http://bautheac.duckdns.org:3838/bautheac/co-movement/}{Internet
Appendix}. Factor construction methods along with \textbf{backwardation}
and \textbf{contango} regimes as well as period definitions are
discussed in Section~\ref{sec-methods} while the results are discussed
in Section~\ref{sec-results} and the commodity assets taxonomy is
detailed in Section~\ref{sec-taxonomy}.}

>>>>

Commodity index swaps, commodity index funds (CIFs), commodity-based
exchange traded funds (ETFs) and notes (ETNs) or commodity linked notes
(CLNs) were amongst the main investment vehicles that allowed
institutional and retail investors to build up commodity exposure in the
early 2000s
\citep{boons_stock_2012, henderson_new_2015, irwin_index_2011, unctad_global_2009, schumann_hunger_2011}.
Most of these products were primarily designed to track prominent
commodity indexes among which the S\&P-GSCI, the most popular.\\
Most of its constituents at the time of writing form the basis of the
broad cross-section of commodities that we consider in this study:
corn-\#2 yellow (XCBT), oats (XCBT), soybean meal (XCBT), soybean oil
(XCBT), soybeans (XCBT), wheat-SRW (XCBT), cattle-feeder (XCME),
cattle-live (XCME), lean hogs (XCME), cocoa (IFUS), coffee-C (IFUS),
cotton-\#2 (IFUS), lumber (XCME), orange juice (IFUS), sugar-\#11
(IFUS), natural gas (XNYM), crude oil-WTI (XNYM), gasoline (XNYM),
heating oil (XNYM), copper (XCEC), gold (XCEC), palladium (XNYM),
platinum (XNYM) and silver (XCEC). Unleaded gasoline, the NYMEX legacy
contract for gasoline, stopped trading in 2006 when it was replaced by
Reformulated Gasoline Blendstock for Oxygen Blending (RBOB) with the two
futures series trading alongside for most of the year. For our NYMEX
gasoline data series, we consider unleaded gasoline up to September
1\textsuperscript{st}, 2006 and RBOB gasoline thereafter, date at which
liquidity for the latter overtook that for the former.\\
We also consider the GB metal complex with aluminium (XLME), copper
(XLME), lead (XLME), nickel-primary (XLME), tin-refined (XLME) and zinc
(XLME). This sample cross-section of assets is further divided into
groups for the purpose of the analysis by country (US, GB), sector
(agriculturals, energy, metals) and by sub-sector (agriculturals:
grains, livestock, softs; energy: gas, petroleum; metals: base,
precious)\footnote{See Table~\ref{tbl-assets-taxonomy} for a detailed
  taxonomy of the commodity assets considered in the study.}.

@@

  attached \href{https://bautheac.shinyapps.io/co-movement/}{Internet

@@

Financialisation was a global phenomenon, through the channel of index
investment it simultaneously influenced the entire spectrum of liquid
commodities, eventually spurring the transformation of commodity markets
from a collection of idiosyncratic markets into a more cohesive
collection akin to an asset class. In this context, interconnectedness
emerges as a crucial aspect that warrants thorough examination.\\
We strive to address this issue by first examining the average returns
pairwise correlations between commodities across the entire
cross-section of the commodity assets considered in the study as well as
by country, sector and sub-sector.
Table~\ref{tbl-correlations-inner-periods} shows the results for the
four periods of interest with the results reported for the entire period
as well as for periods of \textbf{backwardation} and \textbf{contango}.
For the entire sample the average pairwise correlation jumps from 0.07
in the first period to 0.16 over financialisation. This finding
indicates higher interdependence within the full cross-section of
commodities starting with the financialisation period, consistent with
the empirical findings in \citet{fryMcKibbin_evolution_2023} and
\citet{mayer_financialization_2017} as well as with the theoretical
model of \citet{basak_model_2016}. The sector analysis sheds light on
the driving force behind these changes. The US metals see the sharpest
increase in pairwise correlations going from 0.22 over the past period
to 0.56 over financialisation while the three other sectors see a much
smaller percentage increase during financialisation relative to the past
period. These results thus suggest that the US metals sector was the
driving force behind increased average pairwise correlations during
financialisation, a new observation to the best of our knowledge. The
pattern of correlations is very similar across phases of
\textbf{backwardation} and \textbf{contango} for all the sectors and
sub-sectors over the past and financialisation periods. In this context,
the ontological consequences of financialisation are better observed
from an investment perspective as illustrated in
Table~\ref{tbl-correlations-cross-periods} where we observe average
pairwise correlation coefficients between equally weighted portfolios of
US traded commodity assets grouped by sub-sector\footnote{A more
  detailed breakdown of country, sector and subsector average pairwise
  correlations is provided in the attached
  \href{https://bautheac.shinyapps.io/co-movement/}{Internet Appendix}.}.
Although modest in magnitude, these exhibit a telling pattern, mirroring
that shown by the individual commodity returns in
Table~\ref{tbl-stats-regime-tests}, with a substantially higher average
coefficient over \textbf{backwardation} phases in the first period (0.12
vs.~0.03) and the pattern reversed over financilisation with an average
coefficient substantially higher in \textbf{contango} (0.12 vs.~0.02).\\
We break down Table~\ref{tbl-correlations-inner-periods}'s average
correlations by year in Table~\ref{tbl-correlations-inner-years}. For
the full sample of assets, the average pairwise correlations remain in
the 0.06--0.08 range from 1998 through 2003, before suddenly increasing
to 0.12 from 2003 to 2004. This suggests 2004 as a distinct break point
for the onset of financialisation, as noted in several studies
\citep{baker_financialization_2021, tang_index_2012}. We further observe
that the US metals appear to have played a major role in this as the
increase seems driven by the sector which average pairwise correlation
more than doubles (0.49 vs 0.22) over the period.

@@

  \href{https://bautheac.shinyapps.io/co-movement/}{Internet Appendix}.}.
The onset of financialisation leads to a substantial improvement in the
performance of the CHP factor with its average \(R^{2}\), albeit still
small at 5.4\%, almost tripling from the past to the financialisation
period in contrast to the other factors that show weaker dynamics with a
slight decrease and a much smaller increase for the open interest and
term structure factors respectively. Here again the increase in pricing
performance for the CHP factor is driven by the metals sector,
particularly the precious sub-sector with the average \(R^{2}\) for the
group jumping to almost 26\% during financialisation from 1.2\% over the
past period. The corresponding figures of 22\% and 1.1\% for the entire
US metals sector stand in sharp contrast with the agricultural sector,
which exhibits no notable change, and the energy sector, where pricing
performance drops substantially from 9.2\% down to 1.6\%. The
improvement in the CHP factor's pricing performance remarkably comes
predominantly from the long leg (low CHP assets: Keynesian ``normal
backwardation'') of the factor\footnote{A breakdown of the performance
  by factor leg is provided in the attached
  \href{https://bautheac.shinyapps.io/co-movement/}{Internet Appendix}.}
with the corresponding mimicking portfolio's performance itself driven
by the metals, the precious sub-sector in particular. Interestingly,
although the term structure factor itself only sees a very modest
increase, its short leg (low roll-yield assets: Working's backwardation)
shows a sizeable rise in pricing performance, yet again driven the
metals sector. The pattern of results for both factors extends to the GB
traded assets and suggest a convergence of Keneysian and Working's term
structure perspective over the financialisation period.\\

@@

\caption{\label{tbl-stats-regime-tests}This table shows mean returns and
volatilities for the individual commodities and two equally weighted
portfolios of US and GB traded assets for the four periods. For each
period, estimates are computed independently over the whole
period--figure shown--as well as over phases of \textbf{backwardation}
(b) and \textbf{contango} (c)--results in brackets. The table further
shows the results of t-tests of significance for the mean returns
computed over the whole periods and t-tests of difference between these
when computed over the two phases independently along with F-tests of
difference for the corresponding variances. Values significant at the
1\%, 5\% and 10\% level are marked with ***, ** and * respectively.
Detailed results are available in the attached
\href{https://bautheac.shinyapps.io/co-movement/}{Internet Appendix}.
\textbf{Backwardation} and \textbf{contango} regimes as well as period
definitions are discussed in Section~\ref{sec-methods} while the results
are discussed in Section~\ref{sec-results}.}

@@

\caption{\label{tbl-regressions-factors}This table shows the average
time series regression \(R^{2}\) for models where the returns series for
all the commodity assets are independently regressed against the return
series for our mimicking portfolios for risk factors in returns,
including the market, CHP, open interest and term structure factors. The
results are computed independently across countries and commodity
sectors and sub-sectors for the four periods. A breakdown of the
estimates computed independently over the whole periods as well as over
phases of \textbf{backwardation} and \textbf{contango} is provided in
the attached \href{https://bautheac.shinyapps.io/co-movement/}{Internet
Appendix}. Factor construction methods along with \textbf{backwardation}
and \textbf{contango} regimes as well as period definitions are
discussed in Section~\ref{sec-methods} while the results are discussed
in Section~\ref{sec-results} and the commodity assets taxonomy is
detailed in Section~\ref{sec-taxonomy}.}

### 2025-08-19 paper.tex
Deploying to gh-pages from @ bautheac/financialisation-co-movement@fba5d0f4311c2970234a42c3d1515050b6d463c6 🚀

\caption{\label{tbl-stats-regime-tests}This table shows mean returns and
volatilities for the individual commodities as well as for two equally
weighted portfolios of US and GB traded assets across the four periods.
For each period, estimates are computed independently over the whole
period (figure shown) as well as over phases of \textbf{backwardation}
(b) and \textbf{contango} (c) with corresponding comparison results
reported in brackets. The table further shows the results of t-tests of
significance for the mean returns computed over the whole periods as
well as t-tests of difference between these when computed over the two
phases independently along with F-tests of difference for the
corresponding variances. Values significant at the 1\%, 5\% and 10\%
level are marked with ***, ** and * respectively. The estimate values
per regime as well as the p-values for the corresponding difference
tests are available in the attached
\href{http://bautheac.duckdns.org:3838/bautheac/co-movement/}{Internet
Appendix}. \textbf{Backwardation} and \textbf{contango} regimes as well
as period definitions are discussed in Section~\ref{sec-methods} while
the results are discussed in Section~\ref{sec-results}.}

>>>>

\caption{\label{tbl-stats-regime-tests}This table shows mean returns and
volatilities for the individual commodities and two equally weighted
portfolios of US and GB traded assets for the four periods. For each
period, estimates are computed independently over the whole
period--figure shown--as well as over phases of \textbf{backwardation}
(b) and \textbf{contango} (c)--results in brackets. The table further
shows the results of t-tests of significance for the mean returns
computed over the whole periods and t-tests of difference between these
when computed over the two phases independently along with F-tests of
difference for the corresponding variances. Values significant at the
1\%, 5\% and 10\% level are marked with ***, ** and * respectively.
Detailed results are available in the attached
\href{https://bautheac.shinyapps.io/co-movement/}{Internet Appendix}.
\textbf{Backwardation} and \textbf{contango} regimes as well as period
definitions are discussed in Section~\ref{sec-methods} while the results
are discussed in Section~\ref{sec-results}.}
